\documentclass[10pt,oneside]{article}
\usepackage{geometry}

\pagestyle{empty}
\geometry{letterpaper,tmargin=.5in,bmargin=.5in,lmargin=.5in,rmargin=.5in,headheight=0in,headsep=0in,footskip=.3in}

\setlength{\parindent}{0in}
\setlength{\parskip}{0in}
\setlength{\itemsep}{0in}
\setlength{\topsep}{0in}
\setlength{\tabcolsep}{0in}

% Name and contact information
\newcommand{\name}{Andrey Asadchev}
\newcommand{\addr}{201 Spedding Hall, Ames, IA, 50011}
\newcommand{\phone}{(515) 294-4604}
\newcommand{\email}{asadchev@gmail.com}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}
  \fontfamily{ppl} \selectfont
  % Name
  \large \textbf{\name}
  
  \vspace{4pt}

  % Address, phone, email
  \small \addr \\ \phone \\ \email

  \rule{\textwidth}{0.5pt}

  \vspace{6pt}
  % Education ojbective
  \normalsize{
  \textbf{Education objective}
  \small
  \begin{itemize}
  \item{
    \begin{tabular*}{6in}{l@{\extracolsep{\fill}}r}
      Ph.D. in Chemistry with emphasis on Computational Chemistry
    \end{tabular*}
  }
  \end{itemize}

  \vspace{6pt}

  % Education
  \normalsize \textbf{Education}
  \small
  \begin{itemize}
  \item{
    \begin{tabular*}{6in}{l@{\extracolsep{\fill}}r}
      \textbf{Valdosta State University (VSU)} & Valdosta, GA
    \end{tabular*} \\
    \begin{tabular*}{6in}{l@{\extracolsep{\fill}}r}
      B.S., Computer Science Major & May 2005 \\
      B.S., Mathematics Minor & May 2005
    \end{tabular*} \\
  }
  \item{
    \begin{tabular*}{6in}{l@{\extracolsep{\fill}}r}
      \textbf{Iowa State University (ISU)} & Ames, IA
    \end{tabular*} \\
    \begin{tabular*}{6in}{l@{\extracolsep{\fill}}r}
      Ph.D., Physical Chemistry & expected Dec 2011 \\
    \end{tabular*} \\
  }

  \end{itemize}

  \normalsize
  \textbf{Research interests}\\
  \small
  \begin {itemize}
    \item CPU and memory efficient algorithms.
    \item Chemistry and Linear Algebra algorithms on GPU accelerators.
    \item Distributed memory algorithms and communication. 
    \item Two electron integral methods, many-body methods, e.g. MP2, CC.
    \item Domain specific embedded languages, e.g tensor expressions.
    \item Higher-order programming and template  meta-programming.
    \item Symbolic computations and automatic code generation.
  \end {itemize}

  \normalsize
  \textbf{Teaching interests}
  \small
  \begin{itemize}
  \item{Applied Mathematics and Computer Science for Science students}
  \item{Advanced C++ and template meta-programming}
  \end{itemize}
  
  \normalsize
  \textbf{Research experience}
  \small
  \begin{itemize}
  \item{\textbf{Dr. Thomas Manning, Valdosta State Univeristy, 2004-2005}} \\
    Linux Beowulf Cluster \\
    Modeling of C8, C10, C12, and C14 cyclic systems
  \item{\textbf{Dr. Mark Gordon, Iowa State University/Ames Lab, 2005-present}} \\
    Parallel CI methods \\
    GAMESS on IBM Blue Gene and Cray XT4 \\
    Massively parallel MP2 program, Supercomputing 08 \\
    Uncontracted Rysq quadrature on GPU, NCSA 09 \\
    Uncontracted Rys Quadrature Implementation of up to G Functions on Graphical Processing Units, JCTC 10
  \end{itemize}

  \normalsize
  \textbf{Related Work Experience}
  \small
  \begin{itemize}
  \item{
    \textbf{Research Assistant, Valdosta State University, Chemistry Dept, 2003-2004} \\
    Chemistry projects with Dr. Thomas Manning.
  }
  \item{
    \textbf{Teaching Assistant, Valdosta State University, Chemistry Dept, 2004} \\
    General Chemistry.
  }
  \item{
    \textbf{Teaching Assistant, Iowa State University, 2005-2006} \\
    General Chemistry recitations and labs.
  }
  \item
    \textbf{Research Assistant, Iowa State University/Ames Lab, 2006-present} \\
    Quantum Chemistry codes and applications.
  \end{itemize}

  \normalsize
  \textbf{Current projects}
  \small
  \begin{itemize}
  \item C++ implementation of HF, MP2, DFT, CC methods in CPU/GPU environment
  \item CC algorithms with $N^3$ memory overhead
  \item Tensor/matrix expressions
  \item CUDA/Boost interfacing, e.g ublas/CUBLAS bindings.
  \end{itemize}
  
  \normalsize
  \textbf{Relevant computer experience}
  \small
  \begin{itemize}
  \item{C++, C, Python, Fortran, Java}
  \item{Standard Template Library and Boost}
  \item{MPI, ARMCI, OpenMP, CUDA}
  \item{POSIX system programming}
  \item{Mathematica, Matlab/Octave}
  \item{UNIX administration}
  \end{itemize}
 
\newpage

  \normalsize
  \textbf{Presentations and Publications}
  \small
  \begin{itemize}
     \item \textbf{Performance of Electronic Structure Calculations on Blue
       Gene/L and Cray XT4 Computers}\\
       A.Asadchev, Poster, 
       \textit{IEEE/ACM Supercomputing 2008}
    \item \textbf{Performance of Electronic Structure Calculations
      on BG/L and XT4 Computers}\\
       A.Asadchev, B.M.Bode, M.S.Gordon,
       \textit{Journal of Computational and Theoretical Nanoscience,
         2009 6(6)}
    \item \textbf{Uncontracted Rys Quadrature on GPU}\\
      A.Asadchev,
      \textit{Path to Petascale, NCSA, 2009}
    \item \textbf{Accelerating Quantum Chemistry Research using GPUs}\\
      A.Asadchev, J.Felder,
      \textit{GPU Technology Conference, NVIDIA, San Jose, 2009}
     \item \textbf{Uncontracted Rys Quadrature Implementation of up to G
       Functions on Graphical Processing Units}\\
       A.Asadchev, V.Allada, J.Felder, B.M.Bode, M.S.Gordon,
       T.L.Windus,
       \textit{Journal of Chemical Theory and Computation 2010 6(3) }
  \end{itemize}

 
\end{document}

