\documentclass[12pt]{article} \usepackage[margin=1in]{geometry}

\usepackage{graphicx}
\usepackage{color}

%% \usepackage{amsmath}
%% \usepackage{listings}

%% \lstloadlanguages{C++} \lstset{ language=C++, breaklines=true,
%%   keywordstyle=\color{blue}, commentstyle=\color{red} }


\newenvironment{listing}%
               {\begin{table}
                   \begin{tabular}{ p{6in} }
                     \hline}%
               {\end{tabular}%
               \end{table}}


\begin{document}

\title{A New Algorithm for Second Order Perturbation Theory}
\author{Andrey Asadchev \and Mark S. Gordon}
\date{}

\maketitle
% \abstract{}

\section{Conclusions}
As the computing technology changes and matures, the scientific
computing must follow.  The hardware and software which was cutting
edge in the 70's and 80's still dictates how many of the computational
chemistry packages are still implemented today.  However, the
computing technology evolved very quickly since the Fortran 77.
Object oriented programming (OOP), generic programming, standard
libraries, and system standards have become the essential pieces of
most modern  commercial and open-source software, small and large
alike.  To keep up with the improvements in computer science, the
computational chemistry algorithms must be either modernized or
rewritten.  Often, due to software architecture decisions made decades
ago, rewriting is the only viable plan for the future.
Not all of the software needs to be  modernized at once:  the key
pieces such as integral and Hartree-Fock methods can be rewritten
alone and integrated into the existing software, one at a time.
Software modernization also presents a opportunity to improve the
existing algorithms, separate them into modular libraries to encourage
reuse among the scientists, and to plan ahead, given the trends in
computing over the last few decades.


The first algorithm presented was the Hartree-Fock, the reference
method in the most election correlation theories.  The Hartree-Fock
method requires evaluation of the two-electron integrals, which
constitutes the most consuming part.  Unlike other pieces in the
computational chemistry, two-electron integral methods are specific to
the domain and do not receive much attention from outside the field.
The integrals were implemented using Rys Quadrature, one out of
several integral methods.  While algorithmically more complex than
other methods, Rys Quadrature is a general numerically stable method
with low memory footprint, which makes it suitable for implementation
on graphical processing units (GPU).  Once the integral engine was
implemented, the multithreaded Hartree-Fock method naturally followed.
The integral and Hartree-Fock GPU implementation was able to reuse
many key pieces of the CPU algorithm, designed to be fast, extensible,
and flexible through the use of code generator and C++ templates.


One of the most common electron correlation methods is second order
many-body perturbation theory, also known as Moller-Plesset second
order perturbation theory (MP2).  Unlike higher-order treatments,
MP2 is relatively inexpe black-box method  which makes it very
popular.  Hence, the Hartree-Fock implementation was followed by MP2
method.  Like the Hartree-Fock method, the MP2  implementation relies
heavily on the fast integrals.  But unlike the Hartree-Fock, most of
computational work is handled by the de facto standard basic linear algebra
subroutines, BLAS.   The MP2 algorithm implemented is a semi-direct
method, meaning that the partially transformed integrals need to be stored
in the secondary storage, such as disk or distributed memory.   Unlike
the other MP2 algorithms, which are either disk or distributed
memory, the implemented algorithm uses OOP features of C++ to provide
transparent integral storage on either disk or in  distributed memory.

The natural follow-up to MP2 is the coupled cluster (CC) theory.
The couple cluster truncated at singles and doubles excitations, CCSD,
with the perturbative triples correction (T) leads to CCSD(T) method,
often called the gold standard of computational chemistry due its
accuracy.  The CCSD(T)  is very expensive method, both in terms of
computer time and memory.  However, with the lessons learned designing
the MP2 algorithm, a fast CCSD(T) algorithm was developed such that it
could run equally a single workstations and supercomputers.
The key to implementation was optimizing the algorithm in terms of
memory first, I/O overhead second, and concentrating on the
computational efficiency last.  By using several properties of atomic
to molecular basis transformations, several expensive computation and
storage requirements were eliminated from the CCSD algorithm.  And by
using well-known loop optimization technique called blocking, the (T)
algorithm was implemented with very little memory and very little I/O
overhead.

The three algorithms summarized above were prompted by
the need to accommodate the wide array of computational hardware.  In
the process, the algorithms were improved, often drastically.
Implemented in C++, the algorithms and the supporting framework were
built as a stand-alone library, with Fortran bindings.  Connected to
GAMESS, the library was successively integrated with the existing
legacy code.  While not explicitly discussed, the supporting
framework, such as basis set and wavefunction objects, is absolutely
necessary to develop robust flexible modern code.


% \bibliographystyle{unsrt}% (uses file ``plain.bst'')
% \bibliography{references}
% \input{document.bbl}

\end{document}



% LocalWords:  
